---
title: "Introduction to Health Data Science Report"
author: "Melanie Rees-Roberts"
date: "2nd December 2024"
output:
  pdf_document: default
  html_document: default
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Understanding health and wealth in Scottish Council areas.

# Background:

Over the last 100 years, life expectancy in the UK has increased
significantly with advances in living standards and advancing healthcare
(1). However, over the last 10 years this has stalled and even declined
in some instances alongside significant inequalities arising across the
country (2-3). In Scotland, disparities in life expectancy are evident
where the most deprived exhibit approximately 10 years difference in
life expectancy compared to their more affluent neighbourhoods (4). In
recent years, the cost of living crisis has also impacted on peoples
health and wellbeing but the extent to which this has impacted on
different income levels is unknown and thereafter how this links to
health (5-6). Available data in the Scottish Public Health Observatory
may held to explore the extent to which people feel able to manage their
own money and how it may act as a useful proxy measure for financial
stress within a cost of living crisis regardless of income levels.

# Research questions:

1.  What is the variation in health and mental health across different
    Scottish council areas?
2.  How is population income associated with the perception of being
    able to manage money well?
3.  How does the ability to manage money well impact on health and
    mental health in Scotland?

# Packages and libraries:

The following packages were used:

-   install.packages("tidyverse")

-   install.packages("car")

The following libraries were used:

-   library(tidyverse)

-   library(readr)

-   library(tidyr)

-   library(dplyr)

-   library(ggplot2)

-   library(stats)

-   library(car)

```{r - packages, eval=FALSE, include=FALSE}
install.packages("tidyverse")
install.packages("car") 
```

```{r - packages and libraries, include=FALSE}

library(tidyverse)
library(readr)
library(tidyr)
library(dplyr)
library(ggplot2)
library(stats)
library(car)

```

# Data acquisition

In this work, we used the datasets listed below. Full descriptions of
each dataset can be found on the [Indicator definitions and
schedule](https://scotland.shinyapps.io/ScotPHO_profiles_tool/) page of
the Scottish Public Health Observatory website.

-   **Population income deprived** - Number and percentage of total
    population classified as income deprived within SIMD income domain.
    Calculated based on 2011 estimates. Last updated August 2021.
-   **Healthy life expectancy, females** - Healthy life expectancy (HLE)
    at birth for females. Last updated June 2023.
-   **Healthy life expectancy, males** - Healthy life expectancy (HLE)
    at birth for males. Last updated June 2023.
-   **Households managing well financially** - Percentage of households
    managing very or quite well financially these days. Respondents were
    asked "Taking everything together, which of these phrases on this
    card best describes how you and your household are managing
    financially these days?" and the six responses ranged from "manage
    very well" to "are in deep financial trouble". The indicator is the
    percentage who responded either "manage very well" or "manage quite
    well". Last updated September 2024.
-   **Mental wellbeing score** - Mean score on the WEMWBS scale (adults,
    16+ years). WEMWBS stands for Warwick-Edinburgh Mental Wellbeing
    Scale. The questionnaire consists of 14 positively worded items
    designed to assess: positive affect (optimism, cheerfulness,
    relaxation) and satisfying interpersonal relationships and positive
    functioning (energy, clear thinking, self-acceptance, personal
    development, mastery and autonomy). It is scored by summing the
    response to each item answered on a 1 to 5 Likert scale ('none of
    the time', 'rarely', 'some of the time', often', 'all of the time').
    The total score ranges from 14 to 70 with higher scores indicating
    greater wellbeing. The SIMD figures were age-standardised to the
    population of Scotland to aid comparison between the quintiles. Last
    updated December 2023

These data were downloaded from
<https://scotland.shinyapps.io/ScotPHO_profiles_tool/> and imported into
R Studio in a single file.

```{r - read in and tidy healthy life expectancy data, results='hide', message=FALSE}

Data <- read_csv("ScotPHO_data.csv")
glimpse(Data) 

```

# Data cleaning and preparation

Data was manipulated to create tidy datasets as each variable was listed
in a single column with the name of the indicator in the column
'indicator'. For each of the four indicators used, tidy datasets were
created.

### Dataset 1 - Healthy life expectancy.

Create tidy data with healthy life expectancy for males and females.

```{r - tidy healthy life expectancy data}

hle <- Data %>% 
  filter(grepl('Healthy', indicator)) %>% 
  select(area_code,area_name,indicator,measure,upper_confidence_interval,
         lower_confidence_interval) %>% 
  rename(sex = indicator,
        hle = measure) %>% 
  mutate(sex = gsub('Healthy life expectancy, females','0', sex,)) %>% 
  mutate(sex = gsub('Healthy life expectancy, males','1', sex,))    
    
head(hle)

```

The data was mutated to create a single mean healthy life expectancy for males and females combined for each Council area.

```{r - create a single mean hle for both males and females combined}
mean_hle <- hle %>% 
  group_by(area_code) %>% 
  mutate(mean_hle = mean(hle)) 

```

This led to duplicate mean healthy life expectancy values, so the data
was sliced to remove the rows containing sex = 0 and remove the column
sex.

```{r - slice the data to remove duplicate mean_hle and remove sex column}
mean_hle1 <- mean_hle %>% 
  filter(sex != 1) %>% 
  select(area_code, area_name, mean_hle)

head(mean_hle1)

```

### Dataset 2 - Population income deprived

The process was repeated for the remaining datasets.

```{r - tidy population income deprived data}

pid <- Data  %>%  
  drop_na()

pid <- pid %>% 
  select(area_code,numerator,measure) %>% 
  rename(population = numerator,
        pid = measure) 
    
```

### Dataset 3 - Mental health score deprived

```{r - tidy mental health score  data}

mental <- Data %>% 
  filter(grepl('Mental', indicator)) %>% 
  select(area_code,measure) %>% 
  rename(mental_score = measure)
         
```

### Dataset 4 - Households managing financially well

```{r - tidy households managing financially well data}

finance <- Data %>% 
  filter(grepl('Households', indicator)) %>% 
  select(area_code,measure) %>% 
  rename(finance = measure)

```

All indicators were then combined into a single table.

```{r - joining of datasets into a single tibble for analysis}

data_final <- inner_join(mean_hle1, pid, by = 'area_code')

data_final <- inner_join(data_final, mental, by = 'area_code') 

data_final <- inner_join(data_final, finance, by = 'area_code') 

head(data_final)
```

# Data analysis and visualisation

## 1. What is the variation in health and mental health across different Scottish council areas?

To visualise the differences in health and mental health across Scottish
councils we plotted each council area in a column plot with councils
ordered on the x axis in order of increasing % of the population that
were income deprived (i.e. income deprivation increases with the x
axis). On top of this, mental wellbeing scores were plotted using points
showing the mental health scores of these council areas.

*Figure 1: The variation in health and mental health score across
different Scottish council areas.*

```{r}
data_final %>% 
  ggplot() +
  geom_col(aes(x = reorder(area_name, pid), y = mean_hle),
           width=0.75, position=position_dodge(0.5)) +
  geom_point(aes(x=area_name, y = mental_score, colour = "red")) +
  coord_cartesian(ylim = c(40, NA)) +
  ggtitle("          Health and mental wellbeing across Scottish Councils") +
  labs(x="Area name by increasing % of population income deprived", 
       y = "Mean Healthy life expectancy in years (bars)
            Mental Wellbeing Score  (dots)") + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
        legend.position = "none")
```

From these data you can see that there is a general decrease in mean
healthy life expectancy with increasing % of the population that is
income deprived. The Orkney Islands stand out as being an outlier to
this trend with a life expectancy of approximately 10 years higher than
its comparable council areas. Mental wellbeing score, however, does not
exhibit a similar correlation with income deprivation and is generally
consistent across council areas with some variation in individual cases
e.g. Angus having the highest mental wellbeing score and East Lothian
one of the lowest despite similar income deprivation levels.

## 2. How is population income associated with the perception of being able to manage money well?

To understand if income deprivation correlates with the ability to
manage money, we compared % of the population that is deprived with the
ability of the population to manage their money well variable as a
scatter plot with best fit line added.

*Figure 2: Correlating the % of the population that are income deprived
with the ability to manage money well in Scottish council areas*

```{r code for Research Q 2, message=FALSE, warning=FALSE}
  data_final %>% 
  ggplot(aes(x = pid, y = finance)) +
  geom_point() +
  geom_smooth(method = lm) +
    ggtitle("Correlation of percentage population income deprived with ability 
          of population to manage their own money for Scottish Council areas") +
  labs(x="% of population that are income deprived", 
       y = "Ability to manage their own money effectively (score)") + 
  coord_cartesian(ylim = c(20, NA)) 

```

The graph shows a negative correlation indicating that those council
areas where there is a higher proportion of people income deprived also
manage their money less well.

```{r - explore correlation between mean_hle and mental health score, include=FALSE, eval=FALSE}
data_final %>% 
  ggplot(aes(x = mean_hle, y = mental_score)) +
  geom_point() +
  geom_smooth(method = lm) +
  ggtitle("Correlation of healthy life expectancy with mental health score") +
  labs(x="Mean Healthy life expectancy", 
       y = "Mental Health Score") + 
  coord_cartesian(ylim = c(40, NA)) 
```

## 3. How does the ability to manage money well impact on health and mental health in Scotland?

### Predicting mean healthly life expectancy using the % of the population income deprived and ability to manage finances well

First, the data were used to create a multiple linear regression model
of mean healthy life expectancy predicted by population income
deprivation and the ability to manage money well.

```{r - linear models}

q3_Model = lm(formula = mean_hle ~ pid + finance, data = data_final)
summary(q3_Model)

```

It is important to check for the normality of the residuals to
understand the model fit and explore any multi-collinearity. In this
model, our residuals were normal in distribution and the collinearity
for variables pid and finance were below 2.5 indicating no considerable
collinearity present in the model between independent variables.

```{r - check the model parameters,eval=FALSE}

q3_Model_residuals = q3_Model$residuals # Get the model residuals
hist(q3_Model_residuals) # Plot the result
qqnorm(q3_Model_residuals) # Plot the residuals
qqline(q3_Model_residuals)# Plot the Q-Q line

# look at collinearity
vif_Values <- vif(q3_Model)
vif_Values %>% 
barplot(main = "VIF Values", horiz = TRUE, col = "steelblue") #create horizontal
#bar chart to display each VIF value
```

The equation for our model can be found in the summary. For this
regression it is:

mean healthy life expectancy = 66.5 - 0.89x + 0.11y

where x is our population income deprived and y is the ability to manage
own finance variable.

The adjusted *R*-squared value is *r*=0.7347. This indicates a moderate
to borderline strong correlation between the variables. Figure 3 below
visualised the relationship.

*Figure 3: Predicting healthy life expectancy in Scottish council areas
using income deprivation and financial management ability*

```{r - plot the model, message=FALSE, warning=FALSE}

data_final %>% 
ggplot(aes(y=mean_hle,x=pid,colour=finance)) +
  geom_point() +
  stat_smooth(method="lm", se=FALSE) +
  ggtitle("Linear regression model to predict healthy life expectancy in 
  Scottish councils using income deprivation and ability to manage
  money well") +
  labs(x="% of population that are income deprived (pid)", 
       y = "Mean healthy life expectancy (mean_hle)",
       colour = "Ability to manage
  money well (finance)")

```

### Predicting mental wellbeing score in Scottish council areas using the % of the population income deprived and ability to manage finances well

The modelling was repeated to predict mental wellbeing score using the %
of the population income deprived and the ability to manage finances
well.

```{r - repeat to model mental health score}

q3_Model_mental = lm(formula = mental_score ~ pid + finance, data = data_final)
summary(q3_Model_mental)

```

```{r - checking parameters of the model, eval=FALSE}

q3_Model_mental_residuals = q3_Model$residuals # Get the model residuals
hist(q3_Model_mental_residuals) # Plot the result
qqnorm(q3_Model_mental_residuals) # Plot the residuals
qqline(q3_Model_mental_residuals) # Plot the Q-Q line

# look at collinearity
vif_Values2 <- vif(q3_Model_mental)
vif_Values2 %>% 
barplot(main = "VIF Values", horiz = TRUE, col = "steelblue") #create horizontal
#bar chart to display each VIF value
```

The residuals were normal and no concerns around collinearity were
identified.

The equation for this model is:

Mental wellbeing score = 49.1 - 0.23x + 0.03y

where x is our population income deprived and y is the ability to manage
own finance variable.

The adjusted *R*-squared value is *r*=0.5359. This indicates a weak
relationship between the variables. Figure 4 below visualises the
relationship. As demonstrated in the plot for research question 1,
mental health seems to be less associated with income deprivation.

*Figure 4: Predicting mental health score using income deprivation and
financial management ability*

```{r - plot the model for mental health score, message=FALSE, warning=FALSE}

data_final %>% 
ggplot(aes(y=mental_score,x=pid,colour=finance)) +
  geom_point() +
  stat_smooth(method="lm", se=FALSE) +
  ggtitle("Linear regression model to predict mental wellbeing score in 
  Scottish councils using income deprivation and ability to manage
  money well") +
  labs(x="% of population that are income deprived (pid)", 
       y = "Mental wellbeing score (mental_score)",
       colour = "Ability to manage
  money well (finance)")

```

# Report summary

In this report we have explored the variation of healthy life expectancy
and mental wellbeing across the 32 Scottish council areas. There is
variation across Scottish council areas in healthy life expectancy but
minimal or small amounts of variation in mental wellbeing scores. Across
Scottish council areas, a trend emerged showing that those areas with
the least proportion of its population being classified as income
deprived (least deprived) had the highest healthy life expectancy. The
most deprived areas had the lowest healthy life expectancy showing a
considerable inequality in healthy life expectancy of 10 years. The
Orkney Islands stood out as a council area that did not conform to this
trend. Although it was difficult to visualise a similar trend for mental
wellbeing score, some council areas with similar levels of deprivation,
for example Angus and East Lothian, had disparate mental wellbeing of
their populations. Understanding and observing trends in life expectancy
and mental health alongside deprivation and other contextual factors may
support policies to improve the general health of Scottish council
areas. These results suggest a focus on reducing deprivation would
improve healthy life expectancy rather than mental wellbeing.

Given the current cost of living crisis, we then went on to understand
if a variable that measures households ability to manage money well
would correlate with the same areas deprivation level. There was a
negative correlation between the ability of households to manage their
money well with increasing income deprivation. This shows that despite
the context of a cost of living crisis, those areas with higher
proportions of the population being income deprived were still less able
to manage their money well. However, we did see a few council areas that
were outliers, for example where areas of least deprivation had higher
and lower abilities in managing money well.

The data was then used to create models for predicting healthy life
expectancy and mental wellbeing using the percentage of the population
that were income deprived and the ability to manage money well. The
models showed that these predictors were better at estimating healthy
life expectancy (*r* = 0.75) compared to mental wellbeing (*r* = 0.53).
These models could help to inform policy and targets for improving
healthy life expectancy in council areas where deprivation and healthy
life expectancy is low.

Limitations to this analysis included data variables collected and
updated in different years. Therefore, the exact ability of the datasets
to reflect each variable at one specific point in time is not possible
and should be consider when interpreting this report. Furthermore, the
variables used, although, similar in scale result from different methods
of computation whereby mental wellbeing score and ability to manage
money well are likert scores converted to a continuous scale using
percentages. Therefore, given the subjectiveness of answering questions
on a likert scale and differing scales, caution should be taken when
interpreting these variables. Finally, we did not perform statistical
testing to confirm correlations performed or predictor variables in the
models created.

# References

1.  Office for National Statistics. [How has life expectancy changed
    over
    time?](https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/lifeexpectancies/articles/howhaslifeexpectancychangedovertime/2015-09-09).
    9 September 2015. Accessed 2.12.24.
2.  The Health Foundation. [Mortality and life expectancy trends in the
    UK](https://www.health.org.uk/publications/reports/mortality-and-life-expectancy-trends-in-the-uk#:~:text=Since%202011%20improvements%20in%20life%20expectancy%20in%20the,is%20affected%20and%20what%20is%20driving%20current%20trends.).
    November 2019. Accessed 2.12.24.
3.  Department for Health and Social Care, UK government. [Chief Medical
    Officer's Report 2021: health in coastal
    communities](https://www.gov.uk/government/publications/chief-medical-officers-annual-report-2021-health-in-coastal-communities).
    21 July 2021. Accessed on 2.12.24.
4.  National Records of Scotland. [Life expectancy in Scotland
    2020-2022, provisional
    figures](https://www.nrscotland.gov.uk/files/statistics/life-expectancy-in-scotland/20-22/life-expectancy-20-22-report.pdf#:~:text=In%20Scotland%2C%20life%20expectancy%20at%20birth%20for%202020-2022,males%20and%205.7%20weeks%20for%20females%20since%202019-2021.).
    26 September 2023. Accessed 2.12.24.
5.  Institute for Government. [Cost of living
    crisis](https://www.instituteforgovernment.org.uk/explainer/cost-living-crisis).
    7 February 2022. Accessed 2.12.24.
6.  The Health Foundation. [The cost-of-living crisis is a health
    emergency
    too](https://www.health.org.uk/news-and-comment/blogs/the-cost-of-living-crisis-is-a-health-emergency-too#:~).
    7 October 2022. Accessed 2.12.24.
